---
title: "Flux Course 2026: Site Explorer"
subtitle: "Run the SSEM calibration exercise at any FLUXNET site"
author: "Michael Dietze & David Moore"
output: html_document
params:
  site_id: "US-MMS"
  year: 2008
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(compiler)
library(tidyverse)
have_GA <- requireNamespace("GA", quietly = TRUE)  ## GA is optional; Section 4.2 skipped if absent
if (have_GA) library(GA) else message("GA package not found — Section 4.2 (GA calibration) will be skipped.\nInstall with: install.packages('GA')")
source("R/functions.R")
source("R/utils.R")

## Ensemble size — increase to 200–5000 on a fast machine
ne <- 100

## Nominal timestep placeholder; prepare_site.R will auto-detect from the data
timestep <- 1800  # seconds; 1800 = half-hourly, 3600 = hourly

## Pull site and year from knit parameters (or defaults if run interactively)
site_id <- params$site_id
year    <- params$year

cat("Date:",   format(Sys.time(), "%Y-%m-%d %H:%M"), "\n")
cat("Commit:", tryCatch(system("git rev-parse --short HEAD", intern = TRUE),
                        error = function(e) "unknown"), "\n")
cat("Site:", site_id, "| Year:", year, "| ne:", ne, "\n")
```

---

# Site Explorer: `r params$site_id`, `r params$year`

This document lets you run the full SSEM model calibration exercise developed by
Mike Dietze (Boston University) at **any FLUXNET site** — not just the US-NR1
example used during the course. It is self-contained: it downloads data if needed,
builds the prior, runs the model, calibrates it, and compares the result to
upscaled FLUXCOM estimates and CMIP6 Earth system model output where those data
are available.

To run this for a different site or year:

```r
rmarkdown::render("FluxCourseSiteExplorer.Rmd",
                  params = list(site_id = "US-NR1", year = 2015))
```

---

# 1. Download site data

If the processed CSV for this site does not already exist, we download the
FLUXNET FULLSET product using the EcosystemEcologyLab fluxnet package.
Downloading and unzipping can take 5–30 minutes depending on site record length
and network speed. The raw ZIP is saved to `data/{site_id}/raw/` (git-ignored);
the processed HH/HR CSV lands in `data/{site_id}/`.

```{r download-data}
## Look for the half-hourly (HH) or hourly (HR) processed CSV
hh_path <- file.path("data", site_id, paste0(site_id, "_HH.csv"))
hr_path <- file.path("data", site_id, paste0(site_id, "_HR.csv"))

if (file.exists(hh_path) || file.exists(hr_path)) {
  cat("Data already present for", site_id, "-- skipping download.\n")
} else {
  message("Downloading ", site_id, " from FLUXNET shuttle...\n",
          "This may take several minutes for a new site.")
  source("data/download_site.R")
}
```

---

# 2. Prepare site data

`prepare_site.R` reads the downloaded CSV and constructs every R object that the
calibration workflow requires. It handles both HH (30-min) and HR (60-min) time
steps automatically, selects ERA5 reanalysis drivers when available, and creates
a unified `flux$NEE_REF` column that works whether the site product provides
`NEE_VUT_REF` (variable u\* threshold) or `NEE_CUT_REF` (constant u\* threshold).

```{r prepare-data}
## rmarkdown locks the 'params' binding in the global environment for parameterized
## reports. prepare_site.R needs to create its own 'params' data.frame (the SSEM
## model parameter ensemble), so we unlock the binding first and stash the Rmd
## params in a private variable.
.rmd_params <- params
if (bindingIsLocked("params", globalenv())) unlockBinding("params", globalenv())
source("data/prepare_site.R")
## After sourcing, 'params' is the [ne x 9] SSEM parameter ensemble data.frame.
## site_id and year were set in the setup chunk and preserved by prepare_site.R's
## if(!exists()) guards.

## After sourcing, the following objects exist in the global environment:
##   flux        — FLUXNET data frame for the target year
##   date        — POSIXct timestamp vector
##   inputs      — data frame with date, temp (°C), PAR (umol/m2/s)
##   nep         — Net Ecosystem Production = -NEE (umol/m2/s)
##   nep.qc      — QC flag (0 = measured, 1–3 = gap-filled)
##   nep.unc     — joint uncertainty on NEE (umol/m2/s)
##   X, X.orig   — [ne × 3] initial C pool ensemble (Mg/ha)
##   params      — [ne × 9] prior parameter ensemble

## timestep is auto-detected by prepare_site.R; update ts_per_day accordingly
ts_per_day <- round(86400 / timestep)  # number of model steps per 24 h
cat("Timestep:", timestep, "s — aggregating", ts_per_day,
    "steps per day for daily plots.\n")
```

---

# 3. Forward run and pre-calibration assessment

## 3.1 Model structure

The SSEM (Super Simple Ecosystem Model) is a three-pool carbon cycle model.
Its state variables are leaf biomass (X[,1]), structural biomass (X[,2]), and
soil organic matter (X[,3]), all in Mg C ha^-1^. Two drivers — PAR and air
temperature — control photosynthesis and soil respiration respectively.

```{r show-model}
SSEM.orig
```

**Task:** Before running the model forward, look at the equations. What processes
are represented? What is missing? What simplifying assumptions might cause problems
at your site (e.g., water limitation in a semi-arid ecosystem, permafrost dynamics
at high-latitude sites)?

## 3.2 Initial conditions (prior)

```{r plot-ic}
pool.lab <- c("leaf", "wood", "SOM")
for (i in 1:3) {
  hist(X[, i], main = paste(site_id, pool.lab[i], "prior"), xlab = "Mg C/ha")
}
```

## 3.3 Prior parameter distributions

```{r plot-params, fig.asp=1}
plot_params(params)
```

## 3.4 Forward ensemble run (Monte Carlo error propagation)

We run `ne` = `r ne` ensemble members forward through time. Each member samples
a different combination of initial conditions and parameter values, so the spread
of trajectories represents our prior uncertainty about the model's behaviour.

```{r forward-run}
system.time({
  output.ensemble <- SSEM(X     = X.orig,
                          params = params,
                          inputs = inputs,
                          verbose = TRUE)
})
```

*Plot all pools and fluxes*

```{r plot-forecast}
plot_forecast(output.ensemble, sample = 5, timestep = ts_per_day)
```

**Task:** Examine the ensemble trajectories for each pool and flux.

* Are the carbon pools and fluxes the right order of magnitude for your site?
* Do the seasonal cycles have the right shape, timing, and amplitude?
* Are the 95% intervals wide or narrow? Does the model appear over- or under-confident?
* Which variables show the most inter-member spread — what does that tell you about
  where the model is most sensitive to its parameter/initial condition uncertainty?

## 3.5 Assessment against eddy covariance NEP

```{r assessment}
## Extract the ensemble NEP (variable 6) and compute the 95% CI
ci <- apply(output.ensemble[, , 6], 1, quantile, c(0.025, 0.5, 0.975))

## Raw half-hourly (or hourly) time series — colours show QC level
ylim <- range(c(range(ci), range(nep, na.rm = TRUE)))
plot(inputs$date, ci[2, ], ylim = ylim, col = 1, pch = 18, cex = 0.2,
     xlab = "Date", ylab = "NEP (umol/m2/s)",
     main = paste(site_id, year, "— raw timeseries (pre-calibration)"))
ciEnvelope(inputs$date, ci[1, ], ci[3, ], col = col.alpha("lightGrey", 0.5))
points(inputs$date, nep, col = nep.qc + 2, cex = 0.02)
legend("topright", legend = c("model", paste("QC", 0:3)),
       pch = c(18, rep(1, 4)), col = c(1, 2:5))

## Predicted vs observed scatter
plot(ci[2, ], nep, col = nep.qc + 1, cex = 0.2,
     xlab = "Model NEP (umol/m2/s)", ylab = "Obs NEP (umol/m2/s)",
     main = paste(site_id, year, "— predicted vs observed (pre-calibration)"))
abline(0, 1, lty = 2)

## Daily time series
date_daily  <- as_date(inputs$date)
daily.nep   <- tapply(nep, date_daily, mean, na.rm = TRUE)
daily.orig  <- apply(output.ensemble[, , 6], 2,
                     function(x) tapply(x, date_daily, mean, na.rm = TRUE))
daily.orig.ci <- apply(daily.orig, 1, quantile, c(0.025, 0.5, 0.975))

ylim <- range(c(range(daily.orig.ci), range(daily.nep, na.rm = TRUE)))
doy  <- sort(unique(date_daily))
plot(doy, daily.orig.ci[2, ], ylim = ylim, col = 1, pch = 18, cex = 0.5,
     xlab = "Date", ylab = "Daily NEP (umol/m2/s)",
     main = paste(site_id, year, "— daily timeseries (pre-calibration)"))
ciEnvelope(doy, daily.orig.ci[1, ], daily.orig.ci[3, ],
           col = col.alpha("lightGrey", 0.5))
points(doy, daily.nep, col = 2, cex = 0.5)
legend("topright", legend = c("data", "orig"), pch = c(1, 18), col = c(2, 1))

## Daily predicted vs observed
plot(daily.orig.ci[2, ], daily.nep, ylim = ylim, xlim = ylim,
     xlab = "Model daily NEP", ylab = "Obs daily NEP",
     main = "Daily predicted vs observed")
abline(0, 1, col = 2, lty = 2)

## Mean diurnal cycle
## Note: the -5 offset below corrects for UTC-5 (US Eastern/Central sites).
## Adjust for your site's UTC offset if needed.
tod          <- inputs$date - as.POSIXct(date_daily) - 5
nep.diurnal  <- tapply(nep, tod, mean, na.rm = TRUE)
time.of.day  <- as.numeric(names(nep.diurnal))
orig.diurnal <- tapply(ci[2, ], tod, mean, na.rm = TRUE)

ylim <- range(range(orig.diurnal), range(nep.diurnal, na.rm = TRUE))
plot(time.of.day, nep.diurnal, col = 2, ylim = ylim,
     xlab = "Time of day (s since midnight, UTC-5)",
     ylab = "NEP (umol/m2/s)",
     main = paste(site_id, "— mean diurnal curve (pre-calibration)"))
lines(time.of.day, orig.diurnal)
legend("topright", legend = c("data", "orig"), pch = c(1, NA),
       lty = c(NA, 1), col = c(2, 1))

## Functional response: NEP vs air temperature
nbin <- 25
Tair <- inputs$temp
xd   <- seq(min(Tair, na.rm = TRUE), max(Tair, na.rm = TRUE), length = nbin)
xmid <- xd[-length(xd)] + diff(xd)
bin  <- cut(Tair, xd)
Tbar.obs <- tapply(nep,     bin, mean, na.rm = TRUE)
Tbar.mod <- tapply(ci[2, ], bin, mean, na.rm = TRUE)
rng <- range(rbind(Tbar.obs, Tbar.mod), na.rm = TRUE)
plot(xmid, Tbar.obs, col = "red", ylim = rng, type = "l",
     xlab = "Air Temperature (°C)", ylab = "NEP (umol/m2/s)", cex.lab = 1.3,
     main = paste(site_id, "— NEP vs temperature (pre-calibration)"))
lines(xmid, Tbar.mod, col = "black", lwd = 4)
legend("topleft", legend = c("Model", "Data"), lwd = 10, col = 1:2,
       lty = 1, cex = 1.7)

## Functional response: NEP vs PAR
PAR  <- inputs$PAR
xd   <- seq(min(PAR, na.rm = TRUE), max(PAR, na.rm = TRUE), length = nbin)
xmid <- xd[-length(xd)] + diff(xd)
bin  <- cut(PAR, xd)
Pbar.obs <- tapply(nep,     bin, mean, na.rm = TRUE)
Pbar.mod <- tapply(ci[2, ], bin, mean, na.rm = TRUE)
rng <- range(rbind(Pbar.obs, Pbar.mod), na.rm = TRUE)
plot(xmid, Pbar.obs, col = "red", ylim = rng, type = "l",
     xlab = "PAR (umol/m2/s)", ylab = "NEP (umol/m2/s)", cex.lab = 1.3,
     main = paste(site_id, "— NEP vs PAR (pre-calibration)"))
lines(xmid, Pbar.mod, col = "black", lwd = 4)
legend("topleft", legend = c("Model", "Data"), lwd = 10, col = 1:2,
       lty = 1, cex = 1.7)
```

**Task:** What have you learned about the model from comparing it to data at your
site?

* Are there errors that suggest parameter misspecification? Do you have specific
  predictions about which parameters need to move and in which direction?
* Are there features of the plots (e.g., a systematic bias at certain temperatures
  or PAR levels) that suggest model structural errors that calibration cannot fix?
* How do the patterns at your site compare to the US-NR1 example shown in class?
  What might explain the differences?

## 3.6 Quantitative skill scores (pre-calibration)

In addition to visual assessment, quantitative skill metrics help you track
whether calibration is actually improving the model and by how much.

```{r skill-pre}
## Use only directly measured observations (QC = 0) for skill assessment
qaqc <- (nep.qc == 0)
E    <- ci[2, qaqc]   # model expected value
O    <- nep[qaqc]     # observed

## Regression
NEE.pre.fit <- lm(O ~ E)

## Build stats table — leave post row blank for now
stats <- as.data.frame(matrix(NA, 2, 5))
colnames(stats) <- c("RMSE", "Bias", "cor", "slope", "R-squared")
rownames(stats) <- c("pre", "post")

stats["pre", "RMSE"]      <- sqrt(mean((E - O)^2))
stats["pre", "Bias"]      <- mean(E - O)
stats["pre", "cor"]       <- cor(E, O)
stats["pre", "slope"]     <- coef(NEE.pre.fit)[2]
stats["pre", "R-squared"] <- summary(NEE.pre.fit)$r.squared

knitr::kable(stats, digits = 4,
             caption = paste(site_id, year,
                             "— model skill (pre-calibration row filled)"))
```

If you liked model skill scores, you might also like:

* Taylor Diagrams
* Continuous Rank Probability Score (CRPS) and log score
* [EF Activities — Model Assessment](https://github.com/EcoForecast/EF_Activities/blob/d92d8a44ee9d654ccbee569d9be3674821402a5c/Exercise_11_ModelAssessment.Rmd#L105)

---

# 4. Sensitivity analysis

A one-at-a-time (OAT) sensitivity analysis shows how model outputs change as each
parameter is varied while all others are held at their ensemble mean. The output
here focuses on mean NEP and four error metrics (RMSE, bias, correlation).

```{r sensitivity, cache=TRUE}
sa.summary <- nee.sensitivity(ns = 10)
vname <- c("param", "mean", "RMSE", "Bias", "cor")
```

```{r sens-plot, fig.asp=1}
## Sensitivity of mean NEP — slope = elasticity (standardised sensitivity)
sens <- sens_plot(var = 2)

## Ranked elasticity table
knitr::kable(sens[order(abs(sens), decreasing = TRUE)],
             caption = "Ranked elasticities (standardised sensitivities)")
```

**Task:**

* Which parameters are most sensitive at your site? Which are least?
* Which are positively associated with NEP? Negatively? Does that make ecological sense?
* Are any sensitivity curves strongly nonlinear? What does that imply for calibration?

```{r sens-error, fig.asp=1}
## Cost function landscapes for each parameter
sens_plot(var = 3, line = FALSE)  # RMSE (minimise)
sens_plot(var = 4, line = FALSE)  # Bias (want ≈ 0)
sens_plot(var = 5, line = FALSE)  # Correlation (maximise)
```

**Task:**

* For each metric, what score would indicate perfect model performance?
* Which parameters show a clear optimum? Which are flat (insensitive)?
* Do the three metrics agree on which direction parameters need to move?

## 4.1 Likelihood function

The likelihood function formalises the cost function idea: it is the probability
of observing the data given a specific set of model parameters. Here we use a
Normal likelihood, assuming independent Gaussian residuals.

```{r likelihood}
likelihood <- function(param, Obs) {
  param <- data.frame(t(param))
  colnames(param) <- c(colnames(params), "sigma")
  Model <- SSEM(X      = Xbar,
                params = param[, 1:9],
                inputs = inputs[, 2:3])
  lnL <- sum(dnorm(Obs, Model[, , 6], sd = param[1, 10], log = TRUE),
             na.rm = TRUE)
  return(lnL)
}

## Parameter mean vector and ranges for optimisation
theta_bar          <- t(apply(params, 2, mean))
theta_bar[10]      <- stats["pre", "RMSE"]
theta_range        <- apply(params, 2, range)
theta_range        <- cbind(theta_range, sigma = c(0, 10))
Xbar               <- matrix(apply(X.orig, 2, mean), nrow = 1)

## Baseline log-likelihood at prior means (sanity check — should not be NA/Inf)
L0 <- likelihood(param = theta_bar, Obs = nep)
cat("Log-likelihood at prior means:", round(L0, 2), "\n")
```

## 4.2 Genetic Algorithm calibration (optional — slow)

The GA searches for the Maximum Likelihood Estimate (MLE). This can take 10–30
minutes. The result is cached to a site- and year-specific file so re-knitting
is fast.

```{r ga-calibration, cache=TRUE, eval=have_GA}
fit_file <- paste0("fitGA_", site_id, "_", year, ".RData")

if (file.exists(fit_file)) {
  load(fit_file)
  cat("Loaded cached GA result from", fit_file, "\n")
} else {
  message("Running GA calibration for ", site_id, " — this may take 10–30 min.")
  fitted_model <- GA::ga(
    type     = "real-valued",
    fitness  = likelihood,
    lower    = theta_range[1, ],
    upper    = theta_range[2, ],
    Obs      = nep,
    parallel = TRUE,
    popSize  = 30,
    maxiter  = 50,
    run      = 10,
    pmutation = 0.2
  )
  save(fitted_model, file = fit_file)
}

summary(fitted_model)
params_MLE <- as.data.frame(fitted_model@solution)
```

If you liked MLE, you may also like:

* Markov Chain Monte Carlo (MCMC) — JAGS, NIMBLE, BayesianTools
* Hamiltonian Monte Carlo — Stan
* Sobol sensitivity analysis to reduce the parameter dimension before calibration

---

# 5. EKI calibration

Ensemble Kalman Inversion (EKI) is an ensemble-based approach to calibration that
is much faster than MCMC or GA for this model. Instead of evaluating thousands of
parameter proposals sequentially, EKI runs all ensemble members in parallel and
nudges the ensemble mean and covariance toward lower error. We iterate 4 times,
picking the iteration with the lowest RMSE.

```{r eki, cache=TRUE}
output  <- list()
param   <- list(first = params)
RMSE    <- 0
lambda  <- 1.2  # variance inflation factor: keeps the ensemble from collapsing too fast

## Run all ensemble members at the mean initial condition to reduce noise
Xbar <- matrix(apply(X.orig, 2, mean), nrow = ne, ncol = 3, byrow = TRUE)

for (i in 1:4) {
  print(i)
  output[[i]]  <- SSEM(Xbar, param[[i]], inputs)
  RMSE[i]      <- sqrt(mean((output[[i]][, , 6] - nep)^2, na.rm = TRUE))
  param[[i+1]] <- EKI(output[[i]][, , 6], Obs = nep,
                       param[[i]], sigma = RMSE[i] * lambda)
  param[[i+1]] <- impose.contraints(param[[i+1]], params)
}

## Final iteration
output[[i+1]] <- SSEM(Xbar, param[[i+1]], inputs)
RMSE[i+1]     <- sqrt(mean((output[[i+1]][, , 6] - nep)^2, na.rm = TRUE))

cat("RMSE by EKI iteration:", round(RMSE, 4), "\n")

## Select the best parameter set, then run with full IC uncertainty
params_new        <- param[[which.min(RMSE)]]
calibrated.ensemble <- SSEM(X.orig, params_new, inputs)
```

## Updated parameters (prior vs. posterior)

```{r eki-params, fig.asp=1}
plot_params(params_new, hist.params = params)
```

---

# 6. Post-calibration assessment

## 6.1 All pools and fluxes

```{r post-forecast}
plot_forecast(calibrated.ensemble, sample = 5, timestep = ts_per_day)
```

## 6.2 NEP timeseries

```{r post-nep}
ci.new <- apply(calibrated.ensemble[, , 6], 1, quantile, c(0.025, 0.5, 0.975))

## Raw timeseries
ylim <- range(c(range(ci.new), range(nep, na.rm = TRUE)))
plot(inputs$date, ci.new[2, ], ylim = ylim, col = 1, pch = 18, cex = 0.2,
     xlab = "Date", ylab = "NEP (umol/m2/s)",
     main = paste(site_id, year, "— raw timeseries (post-calibration)"))
ciEnvelope(inputs$date, ci.new[1, ], ci.new[3, ],
           col = col.alpha("lightGrey", 0.5))
points(inputs$date, nep, col = nep.qc + 1, cex = 0.02)
legend("topright", legend = c("model", paste("QC", 0:3)),
       pch = c(18, rep(1, 4)), col = c(1, 1:4))

## Predicted vs observed
plot(ci.new[2, ], nep, col = nep.qc + 1, cex = 0.2,
     xlab = "Model NEP (umol/m2/s)", ylab = "Obs NEP (umol/m2/s)",
     main = paste(site_id, year, "— predicted vs observed (post-calibration)"))
abline(0, 1, lty = 2)
```

## 6.3 Daily timeseries

```{r post-daily}
daily.calibrated <- apply(calibrated.ensemble[, , 6], 2,
                          function(x) tapply(x, date_daily, mean, na.rm = TRUE))
daily.cal.ci <- apply(daily.calibrated, 1, quantile, c(0.025, 0.5, 0.975))

ylim <- range(c(range(daily.orig.ci), range(daily.cal.ci),
                range(daily.nep, na.rm = TRUE)))
plot(doy, daily.orig.ci[2, ], ylim = ylim, col = 1, pch = 18, cex = 0.5,
     xlab = "Date", ylab = "Daily NEP (umol/m2/s)",
     main = paste(site_id, year, "— daily timeseries (post-calibration)"))
ciEnvelope(doy, daily.orig.ci[1, ], daily.orig.ci[3, ],
           col = col.alpha("lightGrey", 0.5))
points(doy, daily.cal.ci[2, ], col = 4, pch = 18, cex = 0.5)
ciEnvelope(doy, daily.cal.ci[1, ], daily.cal.ci[3, ],
           col = col.alpha("lightBlue", 0.5))
points(doy, daily.nep, col = 2, cex = 0.5)
legend("topright", legend = c("data", "orig", "calib"),
       pch = c(1, 18, 18), col = c(2, 1, 4))

## Daily predicted vs observed
plot(daily.orig.ci[2, ], daily.nep, ylim = ylim, xlim = ylim,
     xlab = "Model daily NEP", ylab = "Obs daily NEP",
     main = "Daily predicted vs observed (orig = black, calib = blue)")
points(daily.cal.ci[2, ], daily.nep, col = 4)
abline(0, 1, col = 2, lty = 2)
```

## 6.4 Mean diurnal cycle

```{r post-diurnal}
cal.diurnal  <- tapply(ci.new[2, ], tod, mean, na.rm = TRUE)

ylim <- range(c(range(cal.diurnal), range(orig.diurnal),
                range(nep.diurnal, na.rm = TRUE)))
plot(time.of.day, nep.diurnal, col = 2, ylim = ylim,
     xlab = "Time of day (s since midnight, UTC-5)",
     ylab = "NEP (umol/m2/s)",
     main = paste(site_id, "— mean diurnal cycle (pre vs post calibration)"))
lines(time.of.day, orig.diurnal)
lines(time.of.day, cal.diurnal, col = 4)
legend("topright", legend = c("data", "orig", "calib"),
       pch = c(1, NA, NA), lty = c(NA, 1, 1), col = c(2, 1, 4))
```

## 6.5 Skill scores (pre and post)

```{r skill-post}
qaqc  <- (nep.qc == 0)
E_new <- ci.new[2, qaqc]
O     <- nep[qaqc]

NEE.post.fit <- lm(O ~ E_new)

stats["post", "RMSE"]      <- sqrt(mean((E_new - O)^2))
stats["post", "Bias"]      <- mean(E_new - O)
stats["post", "cor"]       <- cor(E_new, O)
stats["post", "slope"]     <- coef(NEE.post.fit)[2]
stats["post", "R-squared"] <- summary(NEE.post.fit)$r.squared

knitr::kable(stats, digits = 4,
             caption = paste(site_id, year, "— model skill (pre and post calibration)"))
```

**Task:**

* How did the model perform after calibration? Did all metrics improve?
* Are there places where calibration improved the daily mean but made specific
  hours of the day worse, or vice versa?
* Does the calibration change your understanding of which processes may be
  structurally misrepresented?

---

# 7. How does your site compare to regional and global estimates?

The tower calibration gives us a well-constrained model for one site and one year.
But how does that picture compare to:

* **FLUXCOM** — a machine-learning upscaling of global eddy covariance flux data
  (Nelson et al. 2024, *Biogeosciences*). FLUXCOM provides monthly GPP, NEE, and
  energy fluxes at 0.5° resolution from 1980–2021.
* **CMIP6 Earth system models** — physically-based global climate-carbon cycle
  models (CESM2, IPSL-CM6A-LR, UKESM1-0-LL). These models simulate terrestrial
  carbon fluxes from first principles at coarse spatial resolution (~1–2°).

Comparing tower data to these products is informative, but **requires caution**:
a single tower footprint (~1 km²) is orders of magnitude smaller than a FLUXCOM or
CMIP6 grid cell (tens of thousands of km²). Heterogeneous land cover, urban areas,
and disturbance history within a grid cell can make direct comparisons misleading.
Nevertheless, persistent disagreements between the tower and the global products
often indicate model biases that matter at large scales.

```{r fluxcom-cmip6}
## ── Unit conversions ─────────────────────────────────────────────────────────
## SSEM GPP/NEP: umol CO2 m-2 s-1 (instantaneous rate)
##   Annual total (gC m-2 yr-1) = sum over timesteps × timestep(s) × 12e-6 gC/umol
##
## FLUXCOM CSV (davidjpmoore fork): gC m-2 day-1 (daily means)
##   Annual total = mean daily rate × 365.25 days yr-1
##
## CMIP6 CSV: kg C m-2 s-1 (instantaneous rate)
##   Annual total = mean monthly rate × 1e3 g/kg × 365.25 × 86400 s/yr
kgC_to_gC_yr <- 1e3 * 365.25 * 24 * 3600   # kg C m-2 s-1 → gC m-2 yr-1
d_to_gC_yr   <- 365.25                       # gC m-2 d-1 → gC m-2 yr-1

## ── Derive annual SSEM GPP and NEP ───────────────────────────────────────────
## SSEM output variable indices: 5 = GPP, 6 = NEP (see SSEM.orig code)
## sum × timestep(s) × 12e-6 gC/umol converts umol m-2 s-1 to annual gC m-2
gpp_pre  <- apply(output.ensemble[, , 5],     2, sum) * timestep * 12e-6
gpp_post <- apply(calibrated.ensemble[, , 5], 2, sum) * timestep * 12e-6
nep_pre  <- apply(output.ensemble[, , 6],     2, sum) * timestep * 12e-6
nep_post <- apply(calibrated.ensemble[, , 6], 2, sum) * timestep * 12e-6
## NEE = -NEP; positive NEE = carbon source (atmospheric sign convention)
nee_pre  <- -nep_pre
nee_post <- -nep_post

ssem_annual <- data.frame(
  source   = c("SSEM pre-cal", "SSEM post-cal"),
  gpp_gC   = c(median(gpp_pre),  median(gpp_post)),
  gpp_lo   = c(quantile(gpp_pre,  0.025), quantile(gpp_post,  0.025)),
  gpp_hi   = c(quantile(gpp_pre,  0.975), quantile(gpp_post,  0.975)),
  nee_gC   = c(median(nee_pre),  median(nee_post)),
  nee_lo   = c(quantile(nee_pre,  0.025), quantile(nee_post,  0.025)),
  nee_hi   = c(quantile(nee_pre,  0.975), quantile(nee_post,  0.975))
)

## ── Look for FLUXCOM and CMIP6 files ─────────────────────────────────────────
## CMIP6 filenames use a lower-case dash-free site token (e.g. "US-MMS" → "usmms")
site_token <- tolower(gsub("-", "", site_id))

fluxcom_path <- file.path("data", "fluxcom",
                           paste0(site_id, "_fluxcom_monthly.csv"))
cmip6_paths  <- c(
  CESM2 = file.path("data", "cmip6", paste0("CESM2_",        site_token, "_monthly.csv")),
  IPSL  = file.path("data", "cmip6", paste0("IPSL-CM6A-LR_", site_token, "_monthly.csv")),
  UKESM = file.path("data", "cmip6", paste0("UKESM1-0-LL_",  site_token, "_monthly.csv"))
)

have_fluxcom <- file.exists(fluxcom_path)
have_cmip6   <- any(file.exists(cmip6_paths))

if (!have_fluxcom && !have_cmip6) {
  message("FLUXCOM and CMIP6 data are pre-loaded for US-MMS, US-NR1, DE-Tha, and DK-Sor.",
          " Open dashboard.html in your browser to explore these comparisons interactively.",
          " To add your site, contact the course instructors.")
} else {

  ## ── Load and aggregate FLUXCOM ────────────────────────────────────────────
  ## FLUXCOM CSV columns: year, month, GPP_gC_m2_d, NEE_gC_m2_d, ET_mm_d, TER_gC_m2_d
  ## Units are already gC m-2 day-1 (daily means); positive NEE = source.
  fluxcom_annual <- NULL
  if (have_fluxcom) {
    fluxcom_raw <- read_csv(fluxcom_path, comment = "#", show_col_types = FALSE) |>
      filter(.data$year == .env$year)
    if (nrow(fluxcom_raw) > 0) {
      fluxcom_annual <- fluxcom_raw |>
        summarise(
          gpp_gC = mean(GPP_gC_m2_d, na.rm = TRUE) * d_to_gC_yr,
          nee_gC = mean(NEE_gC_m2_d, na.rm = TRUE) * d_to_gC_yr
        ) |>
        mutate(source = "FLUXCOM")
    } else {
      message("FLUXCOM file found but no rows match year ", year, ".")
    }
  }

  ## ── Load and aggregate CMIP6 ──────────────────────────────────────────────
  ## CMIP6 CSV columns: year, month, gpp_kgC_m2_s, nbp_kgC_m2_s, ...
  ## nbp (net biome production) ≈ NEP for undisturbed sites; NEE ≈ -nbp.
  cmip6_annual <- NULL
  for (model_name in names(cmip6_paths)) {
    path <- cmip6_paths[model_name]
    if (!file.exists(path)) next
    df <- read_csv(path, show_col_types = FALSE) |>
      filter(.data$year == .env$year)
    if (nrow(df) == 0) next
    cmip6_annual <- bind_rows(
      cmip6_annual,
      df |>
        summarise(
          gpp_gC = mean(gpp_kgC_m2_s, na.rm = TRUE) * kgC_to_gC_yr,
          nee_gC = -mean(nbp_kgC_m2_s, na.rm = TRUE) * kgC_to_gC_yr
        ) |>
        mutate(source = model_name)
    )
  }

  ## ── Combine all sources and plot ──────────────────────────────────────────
  comparison <- bind_rows(ssem_annual, fluxcom_annual, cmip6_annual)
  comparison$source <- factor(comparison$source,
                               levels = unique(comparison$source))

  gpp_plot <- ggplot(comparison, aes(x = source, y = gpp_gC, fill = source)) +
    geom_col(show.legend = FALSE) +
    geom_errorbar(aes(ymin = gpp_lo, ymax = gpp_hi), width = 0.2, na.rm = TRUE) +
    labs(title = paste(site_id, year, "— Annual GPP comparison"),
         x = NULL, y = "GPP (gC m⁻² yr⁻¹)") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
  print(gpp_plot)

  nee_plot <- ggplot(comparison, aes(x = source, y = nee_gC, fill = source)) +
    geom_col(show.legend = FALSE) +
    geom_errorbar(aes(ymin = nee_lo, ymax = nee_hi), width = 0.2, na.rm = TRUE) +
    geom_hline(yintercept = 0, lty = 2) +
    labs(title = paste(site_id, year, "— Annual NEE comparison"),
         subtitle = "Positive = carbon source; negative = carbon sink",
         x = NULL, y = "NEE (gC m⁻² yr⁻¹)") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
  print(nee_plot)

  cat("\nAnnual GPP and NEE summary (gC m-2 yr-1):\n")
  print(comparison[, c("source", "gpp_gC", "nee_gC")])

}  ## end if(have_fluxcom || have_cmip6)
```

**A note on spatial scale.** SSEM is calibrated to a single flux tower footprint
(roughly 0.5–2 km² depending on wind conditions and canopy height). FLUXCOM and
CMIP6 grid cells cover 2,500–40,000 km². Within that area there may be roads,
buildings, different vegetation types, and disturbed patches that are not
represented by the tower. Agreement between the tower model and global products
is informative but never expected to be perfect, and disagreements should be
interpreted carefully before concluding that either is "wrong".

---

# 8. Next steps

Some things you might consider next:

* **Additional data constraints** — sensor-based observations that update
  frequently work particularly well: MODIS LAI, SMAP soil moisture, automated
  chamber soil respiration.
* **Modify or improve the model** — many assumptions in SSEM are known to be
  false (no water limitation, no phenology, fixed allocation). Which one matters
  most at your site?
* **Longer calibration** — do parameters vary by season or by year? Run the
  exercise for multiple years and compare the calibrated parameter distributions.
* **Other sites** — the four pre-loaded sites (US-NR1, US-MMS, DE-Tha, DK-Sor)
  span a range of vegetation types and climates. Find a site that represents your
  own research ecosystem at <https://fluxnet.org/sites/> and run this Rmd for it.

**Interactive exploration** — open `dashboard.html` in your browser to compare
SSEM output side-by-side with FLUXNET observations, FLUXCOM upscaling, and CMIP6
projections for the four pre-loaded sites. No internet connection required.

**To run this Rmd for a different site:**

```r
rmarkdown::render(
  "FluxCourseSiteExplorer.Rmd",
  params = list(site_id = "US-NR1", year = 2015)
)
```

Replace `"US-NR1"` with any valid AmeriFlux or ICOS site ID (e.g., `"DE-Tha"`,
`"DK-Sor"`, `"AU-Tum"`) and set `year` to a year within the site's data record.
A list of all available sites is at <https://fluxnet.org/sites/>.

---

```{r session-info}
sessionInfo()
```
