---
title: "Flux Course 2026: Carbon cycle model-data fusion"
author: "Michael Dietze"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(compiler)
library(tidyverse)
#library(mvtnorm)    ## optional
#library(GA)         ## optional
#library(numDeriv)   ## optional
source("R/functions.R")
source("R/utils.R")

## configuration

#### SET THE ENSEMBLE SIZE
ne = 100 ## should be 200 - 5000, depending on what your computer can handle

timestep = 1800 #seconds

start_date = as.Date("2015-07-01")

outdir = "./" ## output directory for saving files
```

# Super Simple Ecosystem Model

Let's begin by defining our model itself, as well as a number of ancillary 
functions that will be useful in simulation and analysis. The model below is 
very simple but is complex enough to have some chance at capturing observed 
variability. FYI the code for this model, and most of the custom functions below, 
live in R/functions.R.

```{r}
SSEM.orig
```

In terms of inputs, the model has three state variables (X) that are all expressed in
terms of carbon (Mg/ha): Leaf Biomass (X[,1]), Structural Biomass (wood, roots, etc; X[,2]),
and soil organic matter (SOM; X[,3]). The modeled variables are thus the columns of X, while (as we'll see shortly) the different rows represent different replicate ensemble members. To be able to run the model we'll need to define the starting values of these variables (see **Initial Conditions**)

The model also has only two drivers, photosynthetically active radiation (PAR) and air temperature. We pass these into the model as a data frame "inputs" with columns "PAR" and temp" (see **Drivers**). The model parameters are similarly passed in through the data frame "params" and are defined in the **Parameters** section below.

Within the model we first 
estimate LAI from Leaf Biomass and specific leaf area (SLA). Using LAI and light we estimate GPP 
using a simple light use efficiency approach. GPP is then allocated to 
autotrophic respiration (Ra), leaf NPP, and woody NPP based on fixed allocation fractions. These leaf and wood 
biomass pools can then turns over into SOM as litterfall and Coarse Woody Debris / mortality assuming fixed turnover rates. 
SOM is then lost via heterotrophic respiration, which is assumed to be proportional to the size of the SOM pool and to follow a standard Q10 temperature sensitivity.  

**Task:** It's very easy to think of ways that you could make a model more complicated, but not all additional complexity is needed. Remember the goal of modeling is parsimony (i.e., to be as simple as possible but no simpler). Are there ways the current model could be further simplified? If there's a process your gut really wants to add to the model, what's the simplest way you could represent it? Any addition of complexity will always improve model fit to the data you use for calibration, so "fit" isn't a good measure of parsimony -- how would you judge whether an addition of complexity is justified/parsimonious? 


# Initial Conditions

Having defined our model, the next step is to define the ensemble size and 
generate an ensemble estimate of the initial state variables. To do so we'll use
the estimates that are reported in the Ameriflux BADM Meta-data files for the 
site. Since we're only relying on two different estimates of pool size to 
calculate our mean and standard deviation, and neither estimate has a reported 
error, these should be taken as "demonstration only" rather than as 
"Best practices". In a real application one would want to account for the 
sampling error associated with the number of vegetation plots or soil cores 
measured, the measurement error in the soil C and tree DBH, and the allometric 
uncertainty in converting from DBH to leaf and stem biomass. In other words, our
pool sizes are likely a lot less certain than what we take them to be in this 
exercise.

There are a few assumptions in developing the initial conditions
- The mean and standard deviation are only calculated using two observations. 
  As stated above, this is not best practice.
- The wood state is the combination of stems, coarse roots, and fine roots
- The SOM (soil) state is the combination of litter, downed coarse woody debris,
  and soil.
- All values are converted from g/m2 to Mg/ha

```{r}
### Initial State (Mg/ha)
### These specific data points were extracted from the US-NR1 Ameriflux BADM files
### https://ameriflux.lbl.gov/sites/siteinfo/US-NR1
### METADATA: https://ameriflux.lbl.gov/data/badm/badm-standards/ 
### data is also available at https://bit.ly/3POkhq9
#library(readxl)
#badm = readxl::read_xlsx(file.path("AMF_AA-Flx_FLUXNET-BIF_CCBY4_20220606.xlsx")) %>% filter(SITE_ID == "US-NR1")
#write_csv(x=badm,file="data/US-NR1_BADM.csv")
#Bwood = badm %>% filter(VARIABLE == "AG_BIOMASS_TREE") %>% select(DATAVALUE) %>% type_convert() ## unsuccessful crack at automating extraction
Bwood = 14500 * 1e-6 * 10000 ## convert g/m2 -> Mg/ha
Bleaf = 2950 * 0.01 * 4.2/14 ## last term is a correction factor as raw obs generates a LAI that is too high
SOM = c(1.57, 1.58) + c(0.49, 1.39) + c(2.06, 2.59) * 1e-3 * 10000 ## sum up litter, CWD, and soil; change units (US-ME2)
X = as.matrix(c(mean(Bleaf), mean(Bwood), mean(SOM)))
### sample initial condition ensemble members
if(ne > 1){
  X = as.matrix(cbind(
      rnorm(ne, X[1], Bleaf*0.1), ## no variability in data, assume 10%
      rnorm(ne, X[2], Bwood*0.1), ## no variability in data, assume 10%
      rnorm(ne, X[3], sd(SOM))))
}
X.orig = X ## make a copy so that we can run different experiments

## visualize initial condition priors
pool.lab = c("leaf", "wood", "SOM")
for(i in 1:3){
  hist(X[, i], main = pool.lab[i])
}
```

# Drivers

As noted before, this simple model only has two environmental covariates -- the effect of PAR on photosythesis and the effect of air temperture on respiration. We grab these from the gap-filled tower-level data
```{r}
## data downloaded from FLUXNET2015 "fullset" product, subset to 2015 for this analysis
# flux <- read_csv("AMF_US-NR1_FLUXNET_FULLSET_HH_1998-2016_3-5.csv")
# flux <- flux %>% filter(date > as.Date("2015-01-01"))
# write_csv(x=flux,file= "data/US-NR1_Flux2015.csv")
flux <- read_csv("data/US-NR1_Flux2015.csv")
date = strptime(flux$TIMESTAMP_START,format="%Y%m%d%H%M")

inputs <- data.frame(
  date = date,
  temp = flux$TA_ERA, ## gap filled air temperature (Celsius)
  PAR  = flux$SW_IN_ERA / 0.486 ## gap filled PAR, conversion From Campbell and Norman p151
)
plot(inputs$date,inputs$PAR,type='l')
plot(inputs$date,inputs$temp,type='l')
```

# Parameters: Priors & Direct constraints (trait data)

This section defines our initial _a priori_ estimates of the model parameters before we confront the model with the observed eddy covariance data. As you see below, these are defined through a combination of site-specific trait data, literature values, and expert option. In all cases we aimed to define not just a single "default" value for the parameters but also our _a priori_ uncertainty. This prior uncertainty is important for understanding model sensitivity/uncertainty and for constraining the model calibration. For example, if a particular parameter is well known based on trait data, we wouldn't want to give it as much "wiggle room" as parameters that are based on rough guesses.
```{r}
## ancillary data from Ameriflux BADM metadata
SLA = 1e3/c(193.7,205.1,237.7)     ## m2/kg
litterfall = c(215.8)*0.01*3 ## gC/m2/yr->Mg/ha/yr

### initial params
params = list()

## univariate priors: expert opinion
params$alpha = rlnorm(ne, log(0.02), 0.05)     ## light use efficiency
params$Q10 = rnorm(ne, 2.1, 0.5)               ## soil respiration Q10
params$Rbasal = rlnorm(ne, log(0.2), 1.5) / (params$Q10^2.5) ## Soil basal respiration (umol/m2/sec per Mg/ha of SOM)

## multivariate prior on allocation parameters
## assume that NPP is ~50% of GPP on average (Litton et al 2007)
Ra = 0.5                                    
## prior mean on allocation, assume leaf NPP is 31.5% of total (Quaife et al 2008)
alloc = matrix(c(Ra, (1 - 0.315) * (1 - Ra), 0.315 * (1 - Ra)), 1) 
## draw effective sample size to add stochasticity to prior
Neff = matrix(rnbinom(ne, 1000,mu=50), ne)
rdirichlet <- cmpfun(rdirichlet.orig)         ## byte compile to speed up
## prior on fractional allocation to [Ra, wood, leaf] (each draw must sum to 1)
params$falloc = rdirichlet(ne,Neff%*%alloc) 

## Specific leaf area
params$SLA = rnorm(ne, mean(SLA), sd(SLA)) 

## simulate litterfall turnover based on observed 
## litterfall rate and Bleaf prior (initial condition)
if(length(litterfall)>1){
  sdlitterfall = sd(litterfall)
} else {
  sdlitterfall = 0.1*litterfall ## wild assumption
}
lit = rnorm(10000,mean(litterfall),sdlitterfall/sqrt(2))/     
  rnorm(10000,mean(X[,1]),sd(X[,1])/sqrt(2))     ## X1 = leaf biomass 
## draw prior mean and sd; convert turnover per year -> turnover per timestep
lit.mu = rnorm(ne,mean(lit),sd(lit))*timestep/86400/365  
lit.sd = 1/sqrt(rgamma(ne,10,10*var(lit)))*timestep/86400/365
litterfall.param = beta.match(lit.mu,lit.sd^2)
## match moments and draw litterfall prior
params$litterfall = rbeta(ne,litterfall.param$a,litterfall.param$b) 

## draw prior mean based on background tree mortality rate of 1/142 per year (Dietze et al 2011)
mortality.mu = 1/rpois(ne,142)*timestep/86400/365             
## draw prior sd assuming a 50% CV
mortality.sd = rbeta(ne,4,4)*mortality.mu*timestep/86400/365         
## match moments and draw mortality prior
mortality.param = beta.match(mortality.mu,mortality.sd^2)
params$mortality = rbeta(ne,mortality.param$a,mortality.param$b)  

## flatten to a data frame
params = as.data.frame(params)
params[params < 0] = 0
```

### plot parameter values
```{r, fig.asp=1}
plot_params(params)
```

# Prior Predictive Checks: Monte Carlo Error Propagation

Armed with a full set of initial conditions, meteorological drivers, and model prior parameter samples we're now ready to run our model forward and do some pre-calibration validation analyses.

By running the model forward many times as an ensemble that samples the model inputs, rather than just a single best estimate, we're generating a probabilistic prediction. There are many was to propagate uncertainties into such predictions, but this ensemble approach employs what is known as a Monte Carlo method, which basically means that we're approximating a probability distribution with samples from that distribution (in this case, samples of the predictive distribution of future C pools and fluxes).

*Run model forward*
```{r}
system.time({
  output.ensemble = SSEM(X=X.orig,
                          params = params,
                          inputs = inputs, 
                          verbose= TRUE) 
})
```

*Plot timeseries*
Here we'll visualize both the 95% interval estimate, the median (black) and a random sample of 5 ensemble members (red) as examples. These plots focus on the outputs of all model pools, not just those that we have observations for. Comparing these predictions to observations will be our next step. For now, stop and think about the model's predictions -- do they seem reasonable? 

**Task:** Some things to consider when looking at a model's outputs:

* Are the pools and fluxes the right order of magnitude?
* Are there directions trends? Do these seem reasonable or are they too steep or too shallow?
* Are there seasonal cycles? Does the shape make sense? The minimum? The maximum? The seasonal amplitude? 
* Are the interval estimated wide or narrow? Is the model likely to be underconfident (intervals too wide) or overconfident (observations likely to fall outside the interval)?
* With time series, points in time are not independent (e.g., 95% intervals contain 95% of the trajectories). What do we think of the sample of trajectories?

```{r}
plot_forecast(output.ensemble,sample=5,timestep = 48) ## average over each 48 timesteps (=daily)
```

If you like Ensemble Forecasting, you may also like...

* Alternative approaches to error propagation
  * Analytical distributional transformations
  * Analytical Moments
  * Linear Tangent approximations
* Uncertainty partitioning
* Predictability analyses (e.g. error vs lead time)
* Dietze 2017 Prediction in ecology: a first-principles framework. Ecological Applications 27: 2048–2060. DOI: 10.1002/eap.1589

# Assessment

At the half-hourly or hourly resolution, the _QC variable indicates if the corresponding record is a measured value (_QC = 0) or the quality level of the gap-filling that was used for that record (_QC = 1, _QC = 3 worse quality).
```{r}
## model
ci = apply(output.ensemble[, ,6], 1, quantile, c(0.025, 0.5, 0.975)) ## forecast confidence interval
                                                                     ## variable 6 = NEP
## observations
## NEE_REF is created by data/prepare_site.R and holds NEE_VUT_REF where
## available, or NEE_CUT_REF with a printed note for sites where VUT is
## not available (e.g. US-NR1 in the AmeriFlux FLUXNET v1.3 product).
nep     = -flux$NEE_REF
nep.qc  = flux$NEE_REF_QC
nep.unc = flux$NEE_REF_JOINTUNC

## plot raw timeseries
ylim = range(c(range(ci),range(nep,na.rm=TRUE)))
plot(inputs$date,ci[2,],ylim=ylim,col=1,pch=18,cex=0.2)
ciEnvelope(inputs$date, ci[1, ], ci[3, ], col = col.alpha("lightGrey", 0.5))
points(inputs$date,nep,col=nep.qc+2,cex=0.02)
legend("topright",legend=c("model",paste("QC",0:3)),
       pch = c(18,rep(1,4)),
       col=  c(1,2:5))

## raw predicted/observed plot
plot(ci[2,],nep,col=nep.qc+1,cex=0.2)
abline(0,1,lty=2)

## smoothed daily time series
date = as_date(inputs$date)
daily.nep = tapply(nep,date,mean,na.rm=TRUE)
daily.orig = apply(output.ensemble[,,6],2,
                         function(x){tapply(x,date,mean,na.rm=TRUE)})
daily.orig.ci = apply(daily.orig, 1, quantile, c(0.025, 0.5, 0.975))

ylim = range(c(range(daily.orig.ci),range(daily.nep,na.rm=TRUE)))
doy  = sort(unique(date))
plot(doy,daily.orig.ci[2,],ylim=ylim,col=1,pch=18,cex=0.5)
ciEnvelope(doy, daily.orig.ci[1, ], daily.orig.ci[3, ], col = col.alpha("lightGrey", 0.5))
points(doy,daily.nep,col=2,cex=0.5)
legend("topright",legend=c("data","orig"),
       pch = c(18,rep(1,2)),
       col=  c(2,1))

## daily predicted/observed
plot(daily.orig.ci[2,],daily.nep,ylim=ylim,xlim=ylim)
abline(0,1,col=2,lty=2)

## mean diurnal curve
tod = inputs$date - as.POSIXct(date)-5
nep.diurnal = tapply(nep,tod,mean)
time.of.day = as.numeric(names(nep.diurnal))
orig.diurnal = tapply(ci[2,],tod,mean)

ylim=range(range(orig.diurnal),range(nep.diurnal,na.rm=TRUE))
plot(time.of.day,nep.diurnal,col=2,ylim=ylim)
lines(time.of.day,orig.diurnal)
legend("topright",legend=c("data","orig"),
       pch = c(18,rep(1,2)),
       col=  c(2,1))

## binned functional responses
nbin = 25
Tair = inputs$temp
xd = seq(min(Tair),max(Tair),length=nbin)
xmid = xd[-length(xd)] + diff(xd)
bin = cut(Tair,xd)
Tbar.obs = tapply(nep,bin,mean,na.rm=TRUE)
Tbar.mod = tapply(ci[2,],bin,mean,na.rm=TRUE)
rng = range(rbind(Tbar.obs,Tbar.mod))
plot(xmid,Tbar.obs,col="red",ylim=rng,type='l',xlab="Air Temperature (C)",ylab="NEP (umol/m2/s)",cex.lab=1.3)
lines(xmid,Tbar.mod,col="black",lwd=4)
legend("topleft",legend=c("Model","Data"),lwd=10,col=1:2,lty=1,cex=1.7)

PAR = inputs$PAR
xd = seq(min(PAR),max(PAR),length=nbin)
xmid = xd[-length(xd)] + diff(xd)
bin = cut(PAR,xd)
Pbar.obs = tapply(nep,bin,mean,na.rm=TRUE)
Pbar.mod = tapply(ci[2,],bin,mean,na.rm=TRUE)
rng = range(rbind(Pbar.obs,Pbar.mod))
plot(xmid,Pbar.obs,col="red",ylim=rng,type='l',xlab="PAR",ylab="NEP (umol/m2/s)",cex.lab=1.3)
lines(xmid,Pbar.mod,col="black",lwd=4)
legend("topleft",legend=c("Model","Data"),lwd=10,col=1:2,lty=1,cex=1.7)

```

**Task:** What have you learned about the model based on it's comparison to data? Are there errors that you think could be attributed to model parameter misspecification? If so, do you have specific predictions about what parameters will need to be adjusted and in what direction to improve fit? Which figures / patterns suggest this?  Are there any features of the pre-calibration checks that suggest problems with model structure that won't be resolvable by calibration?

Also, in terms of assessing overall model performance it is critical to look at as many pools and fluxes as possible, not just NEE. I recommend the following order of assessment, which can be helpful for isolating the specific processes that may be miscalibrated or misrepresented in the model

1. LAI - should primarily be sensitive to initial conditions and SLA
2. unstressed GPP - should primarily be sensitive to photosythetic parameters. If you get the right GPP with the wrong LAI, you got the wrong GPP
3. stressed GPP - checks stress responses (e.g., moisture deficit), usually done by filtering the observed data for specific covariate conditions
4. NPP, biomass, biomass increment - if GPP is correct, this primarily checks autotrophic respiration, but also allocation and turnover
5. leaf and root litter, CWD - if NPP is correct, this primarily checks turnover and secondarily allocation
6. SOC - should primarily be sensitive to initial conditions
7. Soil respiration - primarily checks Rh. Functional responses can be particularly valuable. If you get Rh right but SOC wrong, you got Rh wrong
8. NEE - puts everything together. In can be hard to assess model errors from NEE alone, but diurnal curves can help separate errors in Re vs GPP
9. Soil Moisture (SM)
10. LE - I often check this much earlier for sites that have a strong carbon/water coupling

If you liked model assessment, you might also like...

* Model selection (AIC, DIC, wAIC, Predictive Loss)
* Model averaging

# Quantitative skill scores

In additionally to visually assessing model performance it can be helpful to calculate quantitative measures of model skill. Such metrics rarely tell you something you didn't know from your visual assessments, but they can be really helpful in determining if the things you are doing to improve the model are moving it in the right direction, and how much room for improvement might exist.

Two important concepts when assessing a model are *accuracy* and *precision*. For the error metrics below, think about which are telling you just about accuracy (how far is the prediction from truth), which are telling you about precison (how narrow or broad is the distribution of predictions), and which provide some mixture of the two.

Finally, in the table below we'll fill in the pre-calibration statistics but will leave a row blank for later adding our post-calibration statistics.


```{r}
## apply QAQC
qaqc = (nep.qc == 0)
E = ci[2,qaqc]           ## model = expected
O = nep[qaqc] ## observed

## Model vs obs regressions
NEE.pre.fit = lm(O ~ E)

## performance stats
stats = as.data.frame(matrix(NA,2,4))
colnames(stats) <- c("RMSE","Bias","cor","slope")
rownames(stats) <- c("pre","post")
stats["pre","RMSE"] = sqrt(mean((E-O)^2))
stats["pre",'Bias'] = mean(E-O)
stats["pre",'cor']  = cor(E,O)
stats["pre",'slope'] = coef(NEE.pre.fit)[2]
stats["pre","R-squared"]=summary(NEE.pre.fit)$r.squared
knitr::kable(stats)
```

If you liked model skill scores, you might also like...

* Taylor Diagrams
* Quantile-based model diagnostics, Continuous Rank Probability Score (CRPS), and log score; see [Validation lab](https://github.com/EcoForecast/EF_Activities/blob/d92d8a44ee9d654ccbee569d9be3674821402a5c/Exercise_11_ModelAssessment.Rmd#L105)

# Sensitivity

This section performs a "one at a time" (OAT) sensitivity analysis. In this case, each parameter is set to its mean value and then, one at a time, one of those parameters are varied along a sequence of values. In this case the function uses the parameter `ns` to determine the number of steps and it varies each parameter along a sequence of quantiles from the minimum to the maximum within the ensemble samples. Because this code has to run separate ensembles for each of the 9 model parameters it may take a few minutes to run depending on the ensemble size. The output counter indicates which parameter is currently being run.
```{r}
sa.summary = nee.sensitivity(ns=10)
## output is an array organized by parameter name, parameter samples, and variable
## variables are in the following order: 
vname = c("param","mean","RMSE","Bias","cor")
```

Rather than looking at entire timeseries of model predictions as a function of each parameter value, here we're looking at just a few key summary statistics. The first is the mean NEP averaged over time. The others are error statistics comparing the model to the data (RMSE, bias, correlation). Let's start by visualizing the first of these (mean NEP), as the traditional goal of sensitivity analysis is to understand how the outputs change as a function of the inputs. In addition to make a grid of plots (one for each parameter), here we're also fitting a regression to these plots as these slopes (dNEP/dParameter) represent an overall numeric summary of sensitivity. Because the units of these slopes vary depending on the units of the output variable and the parameter, we also summarize these in a table of standardized sensivities (a.k.a., elasticities) where a value of 1 indicates that a unit change in the parameter produces a unit change in model output, and the sign indicated the direction of the relationship

```{r,fig.asp=1}
## plot sensitivities and calc linear approx
sens = sens_plot(var=2)

## ranked table of elasticities (i.e., standardized sensitivities)
knitr::kable(sens[order(abs(sens),decreasing = TRUE)]) 
```

**Task:**
* Which parameters are most sensitive? Which are least?
* Which are positively associated with NEE? Which are negative? Does this make sense based on your understanding of the carbon cycle?
* Which responses are linear and which are nonlinear? Are there are anywhere where the slope isn't a useful indicator?

# Model Calibration: Cost Functions & Likelihoods

Before we dive into formally defining model calibration, let's take our sensitivity function and use it to look at how different error metrics vary as a function of the different model parameters


```{r,fig.asp=1}
sens_plot(var=3,line=FALSE) ## RMSE (minimize)
sens_plot(var=4,line=FALSE) ## bias (want as close as possible to 0)
sens_plot(var=5,line=FALSE) ## correlation (maximize)
```
** Tasks:**

* For each metric, what score would minimize model error / maximize skill?
* Which parameters need to move in a specific direction (up or down) to improve model skill? Is this consistent with your prior hypotheses?
* Are there parameters that show some sort of concave function that imply a specific parameter value that would minimize error?
* Do the different error metrics tell the same story about the direction a parameter needs to move?

This example of how error varies as a function of model parameters leads us quite naturally to the key idea of model calibation -- finding the model parameter set that minimizes some metric of model error. More formally, calibration aims to minimize a _cost function_. A simple example of a cost function that would be familar from classic statistics (e.g., regression) would be the minimization of the sum of squared errors, which will be equivalent to the minimization of the RMSE that we saw above.

Here we're going to rely on a very common type of cost function called a _Likelihood_, which is a cost function rooted in probability theory. More specifically, a likelihood function, L=P(data|model), is the probability of observing a specific set of observations conditional on a specific model parameter vector. For example, a traditional linear regression assumes a Normal likelihood with residual errors that are independent and identically distributed

$$P(y|X, \beta, \sigma^2) = N\left(y | X\beta,\sigma^2  \right) = \prod_{i=1}^{n} \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{(y_i - \beta^T x_i)^2}{2\sigma^2}\right)$$

In this activity, we're going to similarly assume a Normal likelihood, but in this case the model predictions are provided by SSEM rather than a linear model

$$P(y|\theta, \sigma^2) = N\left(y | SSEM(\theta,X,inputs) ,\sigma^2  \right) $$
where theta is the vector of model parameters (i.e., params), X remain the initial conditions, and inputs remain the meteorological drivers. 

In the code chunk below we implement this Likelihood function in R. This function takes in a vector of parameters and vector of observed data, runs the model, and then uses the `dnorm` function to calculate the Normal likelihood. Note that in dnorm we use the argument `log=TRUE` to return the log of the likelihood, rather than the raw likelihood. This is traditionally done for reasons of numerical stability. First, it's easier to sum up a bunch of log likelihoods than take the product of a bunch of individual likelihoods. Second, probabilities tend to be small numbers, and if you multiply small numbers together you get smaller numbers. When small numbers get small they can easily become inaccurate, and in the worst case they round of to 0, and once L = 0 no further observations can be multiplied by 0 and change the result. By constrast, on a log scale small numbers become large negative numbers, and the sum of these continue to remain numerically accurate.


```{r}
likelihood <- function(param,Obs){
  ## coorce param to 
  param = data.frame(t(param))
  colnames(param) = c(colnames(params),"sigma")
  print(param)
  
  ## run model
  Model = SSEM(X = Xbar,
               params = param[,1:9],
               inputs = inputs[,2:3]) 
  
  ## calculate likelihood
  lnL = sum(dnorm(Obs,Model[,,6],sd = param[1,10],log=TRUE),na.rm = TRUE)
  return(lnL)
}

## some "default" parameter values for testing
theta_bar = t(apply(params, 2, mean))
theta_bar[10] = stats["pre","RMSE"]
theta_range = apply(params, 2, range)
sigma = c(0,10)
theta_range = cbind(theta_range,sigma)
Xbar        = matrix(apply(X.orig,2,mean),nrow=1)

# lnL at the current parameter mean values
# this test is helpful to ensure the likelihood is implemented correctly and doesn't return NaN, NA, or Inf
# It also establishes another baseline metric for assessing improvement
L0 = likelihood(param=theta_bar,Obs=nep)

```

## Numerical Optimization

If the Likelihood is the probability of observing the data given a specific set of model parameters, this implies that _a parameter value is more likely than another if it is the one for which the data are more probable_. This precept, known as the *Likelihood Principle* leads to the further idea that if one parameter value can be more likely than another, can we find the parameter values that are MOST probable. This is know as a **Maximum Likelihood Estimate (MLE)**.

For simple likelihood functions the MLE can be found using calculus: take the derivative, set it equal to 0, solve for maximum. However for more complex models, and in particular for "black box" models (like the one we wrote) that are implemented in code or external simulators, it is common to rely on numerical optimization techniques to find the MLE. 

### Actual calibration takes a long time to run...
```{r}
fit_file = "fitGA.RData"
if(file.exists(fit_file)){ ## load existing calibration if it exists
  load(fit_file)
} else { 
  fitted_model <- GA::ga(
    type = "real-valued", 
    fitness = likelihood,
    lower = theta_range[1,], 
    upper = theta_range[2,],
    
    # --- Arguments passed on to likelihood ---
    Obs = nep,
    
    # --- SPEED TUNING PARAMETERS FOR 9 DIMENSIONS ---
    parallel = TRUE,
    popSize = 30,       # Small population for lower memory/compute overhead
    maxiter = 50,       # Hard ceiling to prevent long run times
    run = 10,           # Stop early if no improvement seen for 10 generations
    pmutation = 0.2     # Higher mutation keeps it escaping local minima despite low popSize
  )
  
  save(fitted_model,file=fit_file)  ## save output for reuse
}

# Summary of the result
summary(fitted_model)
# Extract the best parameter vector
params_MLE <- as.data.frame(fitted_model@solution)
```

Unfortunately, MLE only returns a single best parameter estimate and additional assumptions must be made, and calculations done, to estimate parameter uncertainty. One way of doing this is using a _bootstrap_ method, such as repeatedly resampling the data and refitting the model parameters, repeatng this once for each ensemble member. Another way is using the second derivatives to estimate a covariance matrix (which is only accurate if your log likelihood is roughly quadratic in shape and the minimum is not at one of the boundaries):

```
# Calculate a numerical Richardson extrapolation of the Hessian matrix (Also slow)
H <- numDeriv::hessian(func = function(x) -likelihood(x,Obs=nep),
                                    x = as.vector(best_parameters))
sigma = solve(H)
params_new = as.data.frame(mvtnorm::rmvnorm(ne,mu,sigma))
colnames(params_new) = c(colnames(params),"sigma")
params_new[params_new<0] = 0
plot_params(params_new)                                    
```

If you liked model calibration, you may also like...

* Markov Chain Monte Carlo (MCMC)
  * JAGS, NIMBLE
  * R's BayesianTools
* Hamiltonian Monte Carlo & STAN
* Using uncertainty analysis (Sobol, Monte Carlo, etc) to reduce the number of parameters considered


# Ensemble Kalman Inversion

One of the challenges with numerical optimization (and more sophisticated Bayesian approaches) is that they become very computationally expensive as the model gets more complex (and slower) or the volume of data gets larger. Two key reasons for this are the number of parameter proposals required (which is often thousands) and the fact that the parameter proposals are sequential.

While a bit less intuitive, there are a number of approaches to calibration that are ensemble based, which allow you to parallize model evaluation, and which often require a smaller total number of model runs. Here we're going to illustrate one of the simplest versions of this, which is known as Ensemble Kalman Inversion (EKI). In the EKI we've implemented, the function takes in the ensemble model predictions, the corresponding data, the current parameter matrix, and an estimate of sigma (residual error). It then returns an updated parameter matrix, where the ensemble parameter mean and (co)variance has been nudged. Hopefully this nudge puts us in a better part of parameter space -- sometimes the nudge is too small or too large, so EKI typically works best when run iteratively. However, unlike numerical optimization, EKI usually only requires 3-5 nudges instead of thousands, and it is more easy to parallelize the ensemble runs.

Because of this, in the code below we repeat the following steps:

1. Run an ensemble of model forward runs
2. Calculate the current RMSE as our estimate of sigma
3. Run EKI to update the parameter matrix (param x ensemble member)
4. Check that the parameters respect our constraints (min, max, falloc sums to 1)

Once we've done this a few times, we'll check the RMSE and pick the iteration with the lowest error.

```{r}
## initalize some storage
output  = list()
param   = list(first = params)
RMSE = 0
lambda = 1.2 ## variance inflation factor

## run every ensemble member at the mean IC to reduce noise
Xbar    = matrix(apply(X.orig,2,mean),nrow=ne,ncol=3,byrow = TRUE)

## EKI iterations
for(i in 1:4){  ## will end up with one more parameter set than upper loop
  print(i)
  output[[i]]   = SSEM(Xbar,param[[i]],inputs)
  RMSE[i]       = sqrt(mean((output[[i]][,,6]-nep)^2))
  param[[i+1]]  = EKI(output[[i]][,,6],Obs = nep,param[[i]],sigma=RMSE[i]*lambda)
  param[[i+1]]  = impose.contraints(param[[i+1]],params)
}

## calc RMSE for final iteration
output[[i+1]] = SSEM(Xbar,param[[i+1]],inputs)
RMSE[i+1]     = sqrt(mean((output[[i+1]][,,6]-nep)^2))

RMSE

## select the parameter set that minimizes RMSE, run again with IC uncertainty
params_new = param[[which.min(RMSE)]]
calibrated.ensemble = SSEM(X.orig,params_new,inputs)
```

Comparing the updated parameter estimates to the original
```{r,fig.asp=1}
plot_params(params_new,hist.params = params)
```



* State Space Models
* Kalman Filter
* Ensemble Kalman Filter
* Particle Filter



# Post-calibration prediction


### Timeseries for all variables
```{r}
plot_forecast(calibrated.ensemble,sample=5,timestep = 48)

## model
ci.new = apply(calibrated.ensemble[, ,6], 1, quantile, c(0.025, 0.5, 0.975)) ## forecast confidence interval
                                                                     ## variable 6 = NEP
## plot timeseries
ylim = range(c(range(ci.new),range(nep,na.rm=TRUE)))
plot(inputs$date,ci.new[2,],ylim=ylim,col=1,pch=18,cex=0.2)
ciEnvelope(inputs$date, ci.new[1, ], ci.new[3, ], col = col.alpha("lightGrey", 0.5))
points(inputs$date,nep,col=nep.qc+1,cex=0.02)
legend("topright",legend=c("model",paste("QC",0:3)),
       pch = c(18,rep(1,4)),
       col=  c(1,1:4))
```

### predicted/observed plot
```{r}
plot(ci.new[2,],nep,col=nep.qc+1,cex=0.2)
abline(0,1,lty=2)
```


### Daily timescale plots
```{r}
## smoothed daily time series
date = as_date(inputs$date)
daily.nep = tapply(nep,date,mean,na.rm=TRUE)

daily.calibrated = apply(calibrated.ensemble[,,6],2,
                         function(x){tapply(x,date,mean,na.rm=TRUE)})
daily.cal.ci = apply(daily.calibrated, 1, quantile, c(0.025, 0.5, 0.975))

daily.orig = apply(output.ensemble[,,6],2,
                         function(x){tapply(x,date,mean,na.rm=TRUE)})
daily.orig.ci = apply(daily.orig, 1, quantile, c(0.025, 0.5, 0.975))

ylim = range(c(range(daily.orig.ci),range(daily.cal.ci),range(daily.nep,na.rm=TRUE)))
doy  = sort(unique(date))
plot(doy,daily.orig.ci[2,],ylim=ylim,col=1,pch=18,cex=0.5)
ciEnvelope(doy, daily.orig.ci[1, ], daily.orig.ci[3, ], col = col.alpha("lightGrey", 0.5))
points(doy,daily.cal.ci[2,],col=4,pch=18,cex=0.5)
ciEnvelope(doy, daily.cal.ci[1, ], daily.cal.ci[3, ], col = col.alpha("lightBlue", 0.5))
points(doy,daily.nep,col=2,cex=0.5)
legend("topright",legend=c("data","orig","calib"),
       pch = c(18,rep(1,3)),
       col=  c(2,1,4))

## daily predicted/observed
plot(daily.orig.ci[2,],daily.nep,ylim=ylim,xlim=ylim)
points(daily.cal.ci[2,],daily.nep,col=4)
abline(0,1,col=2,lty=2)
```

### mean diurnal curve
```{r}
tod = inputs$date - as.POSIXct(date)-5
nep.diurnal = tapply(nep,tod,mean)
time.of.day = as.numeric(names(nep.diurnal))
cal.diurnal = tapply(ci.new[2,],tod,mean)
orig.diurnal = tapply(ci[2,],tod,mean)

ylim=range(c(range(cal.diurnal),range(orig.diurnal),range(nep.diurnal,na.rm=TRUE)))
plot(time.of.day,nep.diurnal,col=2,ylim=ylim)
lines(time.of.day,orig.diurnal)
lines(time.of.day,cal.diurnal,col=4)
legend("topright",legend=c("data","orig","calib"),
       pch = c(18,rep(1,3)),
       col=  c(2,1,4))

```

## Skill scores

```{r}
## apply QAQC
qaqc = (nep.qc == 0)
E = ci.new[2,qaqc]           ## model = expected
O = nep[qaqc] ## observed

## Model vs obs regressions
NEE.post.fit = lm(O ~ E)

## performance stats
stats["post","RMSE"] = sqrt(mean((E-O)^2))
stats["post",'Bias'] = mean(E-O)
stats["post",'cor']  = cor(E,O)
stats["post",'slope'] = coef(NEE.post.fit)[2]
stats["post","R-squared"]=summary(NEE.post.fit)$r.squared
knitr::kable(stats)
```

**Task:** 

* How did the model perform after calibration? Are there places where the model has improved? If so, at what timescales and points in time (e.g. time of the day, time of the year)?
* Are there places where an improved "average" fit actually made the model do worse at specific timescales or points in time?
* Does the calibration shift your understanding of what processes might be misrepresented and need structural improvements?

# Next steps

Some things you might consider next

* Additional sites: Fluxnet provides flux and driver data for lots of tower sites, but it might take a bit more work to specify site-specific initial conditions and parameter priors. 
* Additional data constraints to the Likelihood: Things that work particularly well are sensor-based observations that become available frequently (e.g., MODIS LAI, SMAP, soil respiration)
* Modify/improve the model: Lots of assumptions in SSEM are known to be false!
* Longer calibration (do parameters vary by season? by year?)

